Introduction to genomics and DNA methylation

### 2015-09-17 Rmarkdown/biology_background.Rmd
Closes #19
Closes #20.

Environmental variation has clear influence on certain phenotypes. However, it often can be difficult to determine whether phenotypic variation is driven by genetic varation or environment variation, particularly in humans where genetically similar individuals typically also grow up in similar environments.

The above discussion has been about phenotypic variation in a population. But there is also phenotypic variation within the individual. If that sounds strange, consider the fact that in your body a neuron, a leukocyte (a white blood cell) and a cone cell (a photoreceptor in the retina) all have identical genomes. In fact, all cells in an organism, excluding the gametes, have an identical genome[^somatic_mutations] yet play very different biological roles. This is due to different genes being active in different cells.

[^somatic_mutations]: This ignores somatic mutations.

@@

Conrad Waddington coined the phrase in 1942 as a portmanteau of the words 'epigenesis' and 'genetics' \citep{Waddington:2012bu}. Waddington meant epigenetics as the study of how "processes involved in the mechanism by which the genes of the genotype bring about phenotypic effects" \citep{Waddington:2012bu}.

@@

[^dnam_copying]: In practice this copying is not as faithful as, say, the copying of DNA from the parent to the daughter strand. Furthermore, the faithfulness of this copying will be different in different conditions, such as in a healthy liver cell compared to a cancerous liver cell. Nevertheless, the copying of DNA methylation is faithful enough for most biologists to consider it as a mitotically heritable mark, most of the time.

@@

\subsection{Writers, readers and erasers}\label{sec:writers_reads_and_erasers}

@@

Active demethylation is currently a hot topic in epigenetics research. Multiple mechanisms have been proposed, and it is indeed likely that there are multiple ways to achieve active demethylation. These mechanisms were recently reviewed by \citet{Wu:2014gw}, which I briefly summarise:

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/bisulfite_CT_SNP.pdf}
\caption[Sequence variation confounds methylation calling]{Sequence variation between the reference genome and the sample's genome can result in incorrect inference about the methylation state of the sample's genome. The locus in orange is a cytosine in the reference genome but a thymine in the sample's genome. Because the read is compared against the reference genome, it may incorrectly inferred to be an unmethylated cytosine. The locus in purple is a thymine in the reference genome but an unmethylated cytosine in the sample's genome. Because the read is compared against the reference genome, it may incorrectly inferred to be a thymine.}
\label{fig:bisulfite_CT_SNP}
\end{figure}

The bisulfite-treatment of DNA can introduce biases and other problems. Three examples are _PCR-bias_, _incomplete bisulfite conversion_ and _DNA degradation_. PCR-bias is the difference in amplification efficiency of methylated and unmethylated versions of the same DNA sequence \citep{Warnecke:1997eh}. Incomplete bisulfite-conversion leads to cytosines being incorrectly inferred as 5mC since they cytosines were not converted to uracils by the bisulfite-treatment. DNA degredation occurs because sodium bisulfite damages DNA, resulting in the fragmentation of long molecules. This limits the size of the fragments that can be studied using bisulfite-conversion assays to approximately $500$ bp \citep{Ecker:2010et}

@@

The only commercially available assay in this class is the Pacific Biosciences SMRT technology \citep{Flusberg:2010hra}. It infers the presence of DNA modifications by comparing the time it takes to 'read' the modified form of a base, such as 5mC, to the time it takes to read its unmodified form.

>>>>

Environmental variation has clear influence on certain phenotypes. However, it often can be difficult to determine whether phenotypic variation is driven by genetic variation or environment variation, particularly in humans where genetically similar individuals typically also grow up in similar environments.

The above discussion has been about phenotypic variation in a population. But there is also phenotypic variation within the individual. If that sounds strange, consider the fact that in your body a neuron, a leukocyte (a white blood cell), and a cone cell (a photoreceptor in the retina) all have identical genomes. In fact, all cells in an organism, excluding the gametes, have an identical genome[^somatic_mutations], yet play very different biological roles. This is due to different genes being active in different cells.

[^somatic_mutations]: Even this is a simplification. For example, it ignores somatic mutations (the occurrence of a mutation in the somatic tissue of an organism, resulting in a genetically mosaic individual) and V(D)J recombination (which occurs in lymphocytes and is vital for antibody diversity).

@@

Conrad Waddington coined the phrase in 1942 as a portmanteau of the words 'epigenesis' and 'genetics' \citep{Waddington2012-bq}. Waddington meant epigenetics as the study of how "processes involved in the mechanism by which the genes of the genotype bring about phenotypic effects" \citep{Waddington2012-bq}.

@@

[^dnam_copying]: In practice this copying is not as faithful as, say, the copying of DNA from the parent to the daughter strand. Furthermore, the faithfulness of this copying will be different in different conditions, such as in a healthy liver cell compared to a cancerous liver cell. Nevertheless, the copying of DNA methylation is faithful enough for most biologists to consider it as a mitotically heritable mark, most of the time. The enzymes responsible the replication of DNA methylation, the DNA methyltransferases, are discussed in Section \ref{sec:writers_reads_and_erasers}.

@@

\subsection{Writers, readers, and erasers}\label{sec:writers_reads_and_erasers}

@@

Active demethylation is currently an active area in epigenetics research. Multiple mechanisms have been proposed, and it is indeed likely that there are multiple ways to achieve active demethylation. These mechanisms were recently reviewed by \citet{Wu:2014gw}, which I briefly summarise:

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/bisulfite_CT_SNP.pdf}
\caption[Sequence variation confounds methylation calling]{Sequence variation between the reference genome and the sample's genome can result in incorrect inference about the methylation state of the sample's genome. The locus in orange is a cytosine in the reference genome but a thymine in the sample's genome. Because the read is compared against the reference genome, it may be incorrectly inferred to be an unmethylated cytosine. The locus in purple is a thymine in the reference genome but an unmethylated cytosine in the sample's genome. Because the read is compared against the reference genome, it may be incorrectly inferred to be a thymine.}
\label{fig:bisulfite_CT_SNP}
\end{figure}

The bisulfite-treatment of DNA can introduce biases and other problems \citep{Warnecke2002-zl}. Four examples are _PCR-bias_, _incomplete bisulfite-conversion_, _bisulfite over-conversion_, and _DNA degradation_. PCR-bias is the difference in amplification efficiency of methylated and unmethylated versions of the same DNA sequence \citep{Warnecke:1997eh}. Incomplete bisulfite-conversion leads to cytosines being incorrectly inferred as 5mC since they cytosines were not converted to uracils by the bisulfite-treatment. Conversely, bisulfite over-conversion results in a methylated cytosine incorrectly being inferred to be an unmethylated cytosine, although this is uncommon than incomplete bisulfite-conversion \citep{Warnecke2002-zl}. DNA degredation occurs because sodium bisulfite damages DNA, resulting in the fragmentation of long molecules. This limits the size of the fragments that can be studied using bisulfite-conversion assays to approximately $500$ bp \citep{Ecker:2010et}. The influence of these biases can depend on the experimental setup and their potential effects should be born in mind when interpreting results.

@@

The only commercially available assay in this class is the Pacific Biosciences SMRT technology \citep{Flusberg:2010hra}. It infers the presence of DNA modifications by comparing the time it takes to 'read' the modified form of a base, such as 5mC, to the time it takes to read its unmodified form. This does not require that the DNA is bisulfite-converted prior to sequencing. Because the DNA does not undergo bisulfite-conversion (nor the attendant short fragmentation of the DNA), it is in theory possible to analyse DNA methylation from individual, long DNA molecules using Pacific Biosciences SMRT technology[^pacbio]. However, due to the error rate and cost of SMRT sequencing, it is currently all but unfeasible to study genome-wide DNA methylation in mammalian-sized genomes.

[^pacbio]: While bisulfite-treated DNA could be sequenced using SMRT sequencing, this would eliminate the real advantage of the technology, namely, the long reads generated by this sequencer.

The error rate of Pacific Biosciences SMRT sequencing is currently higher than that of Illumina sequencing. This means it is more difficult to make reliable inferences on DNA methylation from individual reads. Given sufficient sequencing coverage, it would be possible to reliably estimate the average level of methylation at a given cytosine, however, the cost of SMRT sequencing all but prohibits high-coverage sequencing of mammalian-sized genomes.
